%\begin{center}
%\underline{\large\textbf{ACKNOWLEDGEMENT}} 
%\end{center}
%
\addcontentsline{toc}{section}{\bf ACKNOWLEDGEMENT}
{\centerline { {\textbf{ACKNOWLEDGEMENT}}}}
\vspace*{0.75cm}

\indent\hspace{0.05cm} With great pleasure and gratitude, I express my sincere thanks to  Prof.~John H.~Maddocks for his supervision, continuous encouragement, and indispensable feedback.
Along with scientific learning, I hope to remember his entertaining and, most of the times, insightful stories.

I am thankful to all the jury members, Prof.~Pablo Dans, Dr.~Oliver Henrich, and Prof.~Jiri Vanicek, for their discerning and insightful remarks, which have substantially improved this thesis and introduced me to several other relevant scientific questions.

This work is a contribution to the cgDNA family of models, which have been developed by several LCVMM members (and collaborators).
I am thankful to all of them, particularly Alessandro, who developed the cgDNA$+$ model and initially helped me understand various codes.
I cannot imagine a kinder and more patient teacher.
Moreover, I am thankful to Prof.~Wilma Olson, Dr.~Luke Czapla, and Dr.~Helen Lindsay for their contributions and insights in the 5$^\text{th}$ chapter of this thesis.
For running MD simulations, I would like to acknowledge SCITAS, HPC facilities at EPFL and, in particular, Gilles for quickly troubleshooting cluster-related problems and Philippe for other computer-related assistance.
I had the pleasure to directly work and collaborate on scientific as well as non-scientific activities (somewhat more cherishable to me) with other LCVMM members, including Chakri, Daiva, Giulio, Harmeet, Jannes, Rasa, Raushan, and Thomas.
Lastly, Carine's contribution to protecting us from non-trivial administrative work deserves much credit.
I am thankful to her for being the kindest person around.

Even though Covid was the only one to be always out there for me (and everyone), I was blessed with many friends in Lausanne and around the world (connected via Zoom).
Firstly, I am most thankful to Raushan and Lucky for being companions in countless activities, out of which playing cards, swimming, and cooking are closest to my heart.
I am also thankful to Abhishek, Challenger, and Vikranth for many and long online discussions on scientific and non-scientific aspects.
Lastly, I would be remiss in not mentioning Ankit, Anurag, Neeraj, Omkar, Ritesh, and Saloni, whose presence and experiences enriched my life.

I would also like to acknowledge all the Alpine mountains for their abundant beauty and serenity. Weekend hikes in these mountains rejuvenated and filled me with fresh inspiration to work on my research endeavors.

I thank Saumya for being the most amazing person and partner in everything.
Her presence, support, understanding, and countless discussions on science, life, and somehow other remaining things made my life delightful and motivating.
I am glad she was there to help polish the thesis and the final presentation.  \clearpage

Last but not least, I am indebted to my parents for constant encouragement and love throughout my life, which is impossible to express in words.
Finally, a big thanks to my dear sister, Parag, for being the most cheerful and lovely person and for endless mischiefs and arguments. %mischeif

This work was funded by the Swiss National Science Foundation, project 200020\_182184.

